% !TEX root = TMV_Documentation.tex

\section{Changes from version \prevtmvversion\ to \tmvversion}
\label{Changes}

This release is mostly just some minor bug fixes.  Also, since Google Code went under,
TMV is now hosted on GitHub at \url{https://github.com/rmjarvis/tmv}.

Here is a list of the changes from version \prevtmvversion\ to \tmvversion.  
(See \S\ref{History} for changes from previous versions to \prevtmvversion\.)
%Whenever a change is not backward compatible, meaning that code using the previous version might be broken, I mark the item with a $\times$ bullet rather than the usual $\bullet$ to indicate this.  
This release is completely backwards compatible as far as the header files and library are 
concerned.  The library should be link-compatible
with previous versions.

The numbers at the ends of some items below indicate which issue is connected
with the change:

\url{https://github.com/rmjarvis/tmv/issues?q=is\%3Aissue+is\%3Aclosed}

\subsection{Bug Fixes}

\begin{itemize}

\item Removed an extraneous 16 bytes of memory usage from \texttt{SmallVector}. (\#12)
\item Gives an error if PREFIX is the source directory (rather than just 
  failing to work properly). (\#13)
\item Fixed compiler warnings from newer versions of clang++.
\item Changed tmv-version to a simple echo line rather than the complicated
  thing we used to have that stopped working when we switched from svn
  to git. (\#20)
\end{itemize}

\subsection{Build updates}

\begin{itemize}
\item Added support for OpenMP for clang v3.7+.
\item Improved automatic library selection for MKL. (\#15)
\item Fixed broken \texttt{USE\_UNKNOWN\_VARS} feature. (\#16)
\item No longer includes \texttt{\$PREFIX/include} in the include path by default.
  Use scons \texttt{IMPORT\_PREFIX=true} if you want to include it. (\#17)
\item Added \texttt{LINKFLAGS} as an explicit option. (\#21)
\item No longer officially support Visual C++ installation.  I suspect it still
  works, but I no longer include it in my list of tested systems.
\end{itemize}

\subsection{Documatation updates}

\begin{itemize}
\item Documented the sort order for eigenvectors returned by \texttt{Eigen} and singular
  values returned \texttt{SV\_Decompose}. (\#14)
\end{itemize}
