% !TEX root = TMV_Documentation.tex

\section{Matrix decompositions}
\label{Decompositions}

While many matrix decompositions are primarily useful for performing matrix division
(or least-squares pseudo-division), one sometimes wants to perform the decompositions for 
their own sake.  It is possible to get at the underlying decomposition with the various
divider accessor routines like \tt{m.lud()}, \tt{m.qrd()}, etc.  However, this is somewhat
roundabout, and at times inefficient.  So we provide direct ways to perform all of
the various matrix decompositions that are implemented by the TMV code.

In some cases, the input matrix, which we call \tt{A} below, can also serve as at least part of the output as well.  For example, the LU decomposition can be done in place, so that on output, $L$ and $U$ are the lower and upper portions of \tt{A}.  Whenever this is the case, \tt{A} will be listed as a non-\tt{const} reference, and there will be a line after the function such as \tt{L = A.lowerTri()} to indicate the portion or portions of the decomposition that are output in \tt{A}.
In other cases, the input matrix is just used as workspace, and it is junk
on output, in which case, there is no such line following the function.
If the input matrix is listed as \tt{const}, then it won't be changed on output.

Also, all the matrix parameters that you pass to these functions may be views as well.  If the matrix is listed as a \tt{const} reference then you may use the corresponding \tt{Const} view type.  If it is listed as a non-\tt{const} reference, the you may use the corresponding non-\tt{Const} view type.

Sometimes, only certain parts of a decomposition are wanted.  For example,
you might want to know the singular values of a matrix, but not care about
the $U$ and $V$ matrices.  For cases such as this, there are versions
of the decomposition routines which omit certain output parameters.
These routines are generally faster than the versions which include all
output parameters, since they can omit some of the calculations.

None of the decompositions are valid for \tt{T = int} or \tt{complex<int>}.

\subsection[LU decomposition] {LU decomposition \rm (\tt{Matrix}, \tt{BandMatrix})}
\index{LU decomposition}

$A \rightarrow P L U$ where $L$ is lower triangular, 
$U$ is upper triangular, and $P$ is a permutation.

\begin{tmvcode}
void LU_Decompose(Matrix<T>& A, Permutation& P);
L = A.unitLowerTri();
U = A.upperTri();

void LU_Decompose(
      const BandMatrix<T>& A, LowerTriMatrix<T>& L, BandMatrix<T>& U, 
      Permutation& P);
\end{tmvcode}
\index{LU decomposition!Matrix}
\index{LU decomposition!BandMatrix}
\index{Matrix!LU decomposition}
\index{BandMatrix!LU decomposition}
In the second case, \tt{U} must have \tt{U.nhi() = A.nlo()+A.nhi()},
and \tt{L} should be \tt{UnitDiag}.

\subsection[Cholesky decomposition] {Cholesky decomposition \rm (\tt{HermMatrix}, \tt{HermBandMatrix})}
\index{Cholesky decomposition}

$A \rightarrow L L^\dagger$, where $L$ is lower triangular,
and $A$ is hermitian.

\begin{tmvcode}
void CH_Decompose(HermMatrix<T>& A);
L = A.lowerTri();

void CH_Decompose(HermBandMatrix<T>& A);
L = A.lowerBand();
\end{tmvcode}
\index{Cholesky decomposition!SymMatrix}
\index{Cholesky decomposition!SymBandMatrix}
\index{SymMatrix!Cholesky decomposition}
\index{SymBandMatrix!Cholesky decomposition}
If \tt{T} is real, then \tt{A} may also be a \tt{SymMatrix} or \tt{SymBandMatrix} respectively, since these are also hermitian.

If $A$ is found to be not positive definite, a \tt{NonPosDef} exception is thrown.

\subsection[Bunch-Kaufman decomposition] {Bunch-Kaufman decomposition \rm (\tt{HermMatrix}, \tt{SymMatrix})}
\index{Bunch-Kaufman decomposition}

$A \rightarrow P L D L^\dagger P^T$ if $A$ is hermitian,
and $A \rightarrow P L D L^T P^T$ if $A$ is symmetric, where $P$ is a permutation,
$L$ is lower triangular, and $D$ is hermitian or symmetric tridiagonal (respectively).  
In fact, $D$ is even more special than that: it is block diagonal with $1 \times 1$
and $2 \times 2$ blocks,
which means that there are no two consecutive
non-zero elements along the off-diagonal.

\begin{tmvcode}
void LDL_Decompose(SymMatrix<T>& A, SymBandMatrix<T>& D, Permutation& P);
L = A.unitLowerTri();
\end{tmvcode}
\index{Bunch-Kaufman decomposition!SymMatrix}
\index{SymMatrix!Bunch-Kaufman decomposition}

Note: If you are using LAPACK, rather than the native TMV code, then the 
\tt{LDL_Decompose} routine throws a
\tt{tmv::Singular} exception if the matrix is found to be exactly singular.  The LAPACK
documentation says that the decomposition is supposed to finish successfully, but I have
not found that to always be true.  So if LAPACK reports that it has found a singular matrix, TMV
will throw an exception.  The native code will always successfully decompose the matrix.
\index{Exceptions!Singular}
\index{LAPACK!Exceptions from LDL\_Decompose}

\subsection[Tridiagonal LDLt decomposition] {Tridiagonal LDL$^\dagger$ decomposition \rm (\tt{HermBandMatrix}, \tt{SymBandMatrix} with \tt{nlo=1})}
\index{Tridiagonal decomposition}

$A \rightarrow L D L^\dagger$ or $A \rightarrow L D L^T$.
where this time $D$ is a regular diagonal matrix and $L$ is
a lower band matrix with a single subdiagonal and all 1's on the diagonal.

It turns out
that the Bunch-Kaufman algorithm on banded matrices tends to expand the band structure
without limit because of the pivoting involved, so it is not practical.
However, with tridiagonal matrices, it is often possible to perform the 
decomposition without pivoting.  There is then no growth of the band structure,
but it is not as stable for singular or nearly singular matrices.
If an exact zero is found on the diagonal along the way
\tt{tmv::NonPosDef} is thrown.\footnote{
Note, however, that if $A$ is complex, symmetric - i.e. not hermitian -
then this doesn't actually mean that $A$ is not positive definite (since such a 
quality is only defined for hermitian matrices).  Furthermore, 
hermitian matrices that are not positive definite will probably be decomposed successfully
without throwing, resulting in D having negative values.

Also, the LAPACK implementation throws an exception for matrices that the native code
successfully decomposes.  It throws for hermitian matrices whenever they are 
not positive definite, whereas the native code succeeds for many indefinite matrices.
}
\index{Exceptions!NonPosDef}
\index{LAPACK!Exceptions from LDL\_Decompose}

\begin{tmvcode}
void LDL_Decompose(HermBandMatrix<T>& A);
void LDL_Decompose(SymBandMatrix<T>& A);
L = A.lowerBand();
L.diag().setAllTo(T(1));
D = DiagMatrixViewOf(A.diag());
\end{tmvcode}
\index{Tridiagonal LDL decomposition!SymBandMatrix}
\index{SymBandMatrix!Tri-diagonal LDL decomposition}

\subsection[QR decomposition] {QR decomposition \rm (\tt{Matrix}, \tt{BandMatrix})}
\index{QR decomposition}

$A \rightarrow Q R$, where $Q$ is column-unitary 
(i.e. $Q^\dagger Q = I$), $R$ is upper triangular, and $A$ is either square or 
has more rows than columns.

\begin{tmvcode}
void QR_Decompose(Matrix<T>& A, UpperTriMatrix<T>& R);
Q = A;

void QR_Decompose(const BandMatrix<T>& A, Matrix<T>& Q, BandMatrix<T>& R);
\end{tmvcode}
\index{QR decomposition!Matrix}
\index{Matrix!QR decomposition}
\index{QR decomposition!BandMatrix}
\index{BandMatrix!QR decomposition}
In the second case, \tt{R} must have \tt{R.nhi()} $>=$ \tt{A.nlo()+A.nhi()}.

If you only need $R$, the following versions are faster, since they do 
not fully calculate $Q$.
\begin{tmvcode}
void QR_Decompose(Matrix<T>& A);
R = A.upperTri();

void QR_Decompose(const BandMatrix<T>& A, BandMatrix<T>& R);
\end{tmvcode}

\subsection[QRP decomposition] {QRP decomposition \rm (\tt{Matrix})}
\index{QRP decomposition}

$A \rightarrow Q R P$, where $Q$ is column-unitary 
(i.e. $Q^\dagger Q = I$), $R$ is upper triangular, $P$ is a permutation, 
and $A$ is either square or has more rows than columns.

\begin{tmvcode}
void QRP_Decompose(
      Matrix<T>& A, UpperTriMatrix<T>& R, Permutation& P, 
      bool strict=false);
Q = A;
\end{tmvcode}
\index{QRP decomposition!Matrix}
\index{Matrix!QRP decomposition}

As discussed in \S\ref{Matrix_Division_Decompositions}, 
there are two slightly different algorithms for doing a QRP decomposition.  
If \tt{strict} is \tt{true}, then the diagonal elements
of $R$ be strictly decreasing (in absolute value) from upper-left to lower-right\footnote{
\index{QRP decomposition!LAPACK ?geqp3}
\index{LAPACK!Problems with QRP decomposition}
If you are using a LAPACK library, you might find that the output $R$ diagonal
is not always strictly decreasing, although it will usually be close.  If strictly monotonic
diagonal elements are important for you, you can use the native TMV algorithm instead
by compiling with the SCons option \tt{USE\_GEQP3=false}. (\tt{geqp3} is the name of the LAPACK 
function that does the QRP decomposition.)}.

If \tt{strict} is \tt{false} (or omitted) however, then the diagonal elements of $R$ will not
be strictly decreasing.  Rather, there will be no diagonal element
of $R$ below and to the right of one which is more than a factor of $\epsilon^{1/4}$ 
smaller in absolute value, where 
$\epsilon$ is the machine precision.  This restriction
is almost always sufficient to make the decomposition useful for singular or nearly
singular matrices, and it is much faster than the strict algorithm for most matrices.

If you only need $R$, the following versions is faster, since it does
not fully calculate $Q$.
\begin{tmvcode}
void QRP_Decompose(Matrix<T>& A, bool strict=false);
R = A.upperTri();
\end{tmvcode}

\subsection[Singular value decomposition] {Singular value decomposition \rm (\tt{Matrix}, \tt{BandMatrix}, \tt{SymMatrix}, \tt{HermMatrix},\\ \tt{SymBandMatrix}, \tt{HermBandMatrix})}
\index{Singular value decomposition}

$A \rightarrow U S V^\dagger$,
where $U$ is column-unitary (i.e. $U^\dagger U = I$),
$S$ is real diagonal, $V$ is square unitary, and $A$ is either square or 
has more rows than columns.
Note that the functions below do not deal with $V$ directly, they instead return $V^\dagger$
in the parameter \tt{Vt}.  So on output, \tt{U*S*Vt} will equal the original matrix \tt{A}.  I find this to usually be more convenient, but if you prefer dealing with $V$ directly, you can pass
\tt{V.adjoint()} as the parameter to these functions.

\begin{tmvcode}
void SV_Decompose(Matrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);
U = A;
V = Vt.adjoint();

void SV_Decompose(
      const SymMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S, 
      Matrix<T>& Vt);
V = Vt.adjoint();

void SV_Decompose(
      const HermMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S, 
      Matrix<T>& Vt);
V = Vt.adjoint();

void SV_Decompose(
      const BandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S, 
      Matrix<T>& Vt);
V = Vt.adjoint();

void SV_Decompose(
      const SymBandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S, 
      Matrix<T>& Vt);
V = Vt.adjoint();

void SV_Decompose(
      const HermBandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S, 
      Matrix<T>& Vt);
V = Vt.adjoint();
\end{tmvcode}
\index{Singular value decomposition!Matrix}
\index{Singular value decomposition!BandMatrix}
\index{Singular value decomposition!SymMatrix}
\index{Singular value decomposition!SymBandMatrix}
\index{Matrix!Singular value decomposition}
\index{BandMatrix!Singular value decomposition}
\index{SymMatrix!Singular value decomposition}
\index{SymBandMatrix!Singular value decomposition}

The input $A$ matrix must not have more columns than rows.  If you want
to calculate the SVD of such a matrix, you should decompose $A^T$ instead:
\begin{tmvcode}
tmv::Matrix<double> A(nrows,ncols); // ncols > nrows
// A = ... 
tmv::Matrix<double> Vt = A;
tmv::DiagMatrix<double> S(nrows);
tmv::Matrix<double> U(nrows,nrows);
SV_Decompose(Vt.transpose(),S.view(),U.transpose());
// Now A = U * S * Vt
\end{tmvcode}

If you only need $S$, or $S$ and $V$, or $S$ and $U$, the following 
versions are faster, since they do
not fully calculate the omitted matrices.  
\begin{tmvcode}
void SV_Decompose(
      Matrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt, bool StoreU);
// With StoreU=false, then U != A

void SV_Decompose(Matrix<T>& A, DiagMatrix<RT>& S, bool StoreU);
if (StoreU) U = A;

void SV_Decompose(SymMatrix<T>& A, DiagMatrix<RT>& S);

void SV_Decompose(HermMatrix<T>& A, DiagMatrix<RT>& S);
      
void SV_Decompose(const SymMatrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);
      
void SV_Decompose(
      const HermMatrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);
      
void SV_Decompose(const SymMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S);
      
void SV_Decompose(
      const HermMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S);
      
void SV_Decompose(const BandMatrix<T>& A, DiagMatrix<RT>& S);

void SV_Decompose(
      const BandMatrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);
      
void SV_Decompose(
      const BandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S);
      
void SV_Decompose(const SymBandMatrix<T>& A, DiagMatrix<RT>& S);
      
void SV_Decompose(const HermBandMatrix<T>& A, DiagMatrix<RT>& S);
      
void SV_Decompose(
      const SymBandMatrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);

void SV_Decompose(
      const HermBandMatrix<T>& A, DiagMatrix<RT>& S, Matrix<T>& Vt);

void SV_Decompose(
      const SymBandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S);

void SV_Decompose(
      const HermBandMatrix<T>& A, Matrix<T>& U, DiagMatrix<RT>& S);
\end{tmvcode}

On output, the singular values \tt{S} will be in decreasing order of absolute value.
This is the typical order one would want for principal value decompositions,
where only the largest several singular values (aka principal values) would be 
used.  Note that this is different from the order returned by \tt{Eigen}
(cf. \S\ref{Eigenvalues}).  Other than this detail and the fact that \tt{S}
is packaged as a \tt{DiagMatrix} rather than a \tt{Vector}, the result of these two
functions for Hermitian matrices is identical.

\subsection[Polar decomposition] {Polar decomposition \rm (\tt{Matrix}, \tt{BandMatrix})}
\index{Polar decomposition}

$A \rightarrow U P$ where $U$ is unitary and $P$ is positive definite hermitian.

This is similar to polar form of a complex number: $z = r e^{i \theta}$.
In the matrix version, $P$ acts as $r$, being in some sense the ``magnitude'' 
of the matrix.  And $U$ acts as $e^{i \theta}$, being a generalized rotation.

\begin{tmvcode}
void Polar_Decompose(Matrix<T>& A, SymMatrix<T>& P);
U = A;

void Polar_Decompose(
      const BandMatrix<T>& A, Matrix<T>& U, SymMatrix<T>& P);
\end{tmvcode}
\index{Polar decomposition!Matrix}
\index{Polar decomposition!BandMatrix}
\index{Matrix!Polar decomposition}
\index{BandMatrix!Polar decomposition}
If \tt{T} is real, then \tt{P} may also be a \tt{SymMatrix}, since it is also hermitian.

\subsection[Matrix square root] {Matrix square root \rm (\tt{HermMatrix}, \tt{HermBandMatrix})}

$A \rightarrow S S$, where $A$ and $S$ are each positive definite hermitian matrices.

\begin{tmvcode}
void SquareRoot(HermMatrix<T>& A);
S = A;

void SquareRoot(const HermBandMatrix<T>& A, SymMatrix<T>& S);
\end{tmvcode}
\index{SymMatrix!square root}
\index{SymBandMatrix!square root}
\index{square root!SymMatrix}
\index{square root!SymBandMatrix}
If \tt{T} is real, then \tt{A} and \tt{S} may also be a \tt{SymMatrix} or \tt{SymBandMatrix} as appropriate, since these are also hermitian.

If $A$ is found to be not positive definite, a \tt{NonPosDef} exception is thrown.
\index{Exceptions!NonPosDef}

\subsection{Update a QR decomposition}
\index{QR decomposition!Update}
\label{QRUpdate}

One reason that it can be useful to create and deal with the QR decomposition directly,
rather than 
just relying on the division routines is the possibility
of updating or ``downdating'' the resulting $R$ matrix.  

If you are doing a 
least-square fit to a large number of linear equations, you can write the system as
a matrix equation: 
$A x = b$, where $A$ is a matrix with more rows than columns, and you are seeking,
not an exact solution for $x$, but rather the value of $x$ which minimizes
$||b-Ax||_2$.  See \S\ref{Matrix_Division_Leastsquare} for a more in-depth discussion of this topic.

It may be the case that you
have more rows (i.e. constraints) than would allow the entire matrix to fit in memory.  
In this case it may be tempting to use the so-called normal equation instead: 
\begin{align*}
A^\dagger A x &= A^\dagger b \\
x & = (A^\dagger A)^{-1} A^\dagger b
\end{align*}
This equation theoretically gives the same 
solution as using the QR decomposition on the original design matrix.
However, it can be shown that the condition of $A^\dagger A$ is the 
\underline{square} of the condition of $A$.  Since larger condition values
lead to larger numerical instabilities and round-off problems, a mildly
ill-conditioned matrix is made much worse by this procedure.

When all of $A$ fits in memory, the better solution is to use the QR decomposition, $A = QR$,
to calculate $x$.
\begin{align*}
Q R x &= b \\
x &= R^{-1} Q^\dagger b
\end{align*}
In fact, this is the usual behind-the-scenes procedure when you write \tt{x = b/A} in TMV.
But if $A$ is too large to fit in memory, then so is $Q$.

A compromise solution, which is not quite as good as doing the full QR decomposition,
but is better than using the normal equation, is to just calculate the $R$ of the
QR decomposition, and not $Q$.  Then:
\begin{align*}
A^\dagger A x &= A^\dagger b \\
R^\dagger Q^\dagger Q R x = R^\dagger R x &= A^\dagger b \\
x &= R^{-1} (R^\dagger)^{-1} A^\dagger b
\end{align*}

Calculating $R$ directly from $A$ is numerically much more stable than 
calculating it through, say, a Cholesky decomposition of $A^\dagger A$.
So this method produces a more accurate answer for $x$ than the normal equation does.

But how can $R$ be calculated if we cannot fit all of $A$ into memory at once?

First, we point out a characteristic of unitary matrices that the product 
of two or more of them is also unitary.  
This implies that if we can calculate
something like: $A = Q_0 Q_1 Q_2 ... Q_n R$, then this is the $R$ that we want.

So, consider breaking $A$ into a submatrix, $A_0$, which can fit into memory, 
plus the remainder, $A_1$, which may or may not.
\begin{equation*}
A = \left(\begin{array}{c}A_0 \\A_1\end{array}\right)
\end{equation*}
First perform a QR decomposition of $A_0 = Q_0 R_0$.  Then we have:
\begin{align*}
A &= \left(\begin{array}{c}Q_0 R_0 \\ A_1 \end{array}\right) \\
&= \left(\begin{array}{cc}Q_0 & 0 \\ 0 & 1\end{array}\right) 
      \left(\begin{array}{c}R_0 \\ A_1 \end{array}\right) \\
&\equiv Q_0^\prime A_1^\prime
\end{align*}

Assuming that $A_0$ has more rows than columns, 
then $A_1^\prime$ has fewer rows than
the original matrix $A$.  So we can iterate this process until the 
resulting matrix can fit in memory, and we can perform the final QR update
to get the final value of $R$.

For the numerical reasons mentioned above, 
the fewer such iterations you do, the better.  So you should try to include as many
rows of the matrix $A$ as possible in each step, given the amount of memory
available.

The solution equation, written above, also needs the quantity $A^\dagger b$, which
can be accumulated in the same blocks:
\begin{equation*}
A^\dagger b = A_0^\dagger b_0 + A_1^\dagger b_1 + ....
\end{equation*}
This, combined with the calculation of $R$, allows us to determine $x$ using the above formula.

The TMV library includes a command which does the update step of the above procedure
directly, which is slightly more efficient than explicitly forming the $A_k^\prime$ matrices.
The commands is
\begin{tmvcode}
void QR_Update(UpperTriMatrix<T>& R, Matrix<T>& X)
\end{tmvcode}
which updates the value of $R$ such that $R_{\rm out}^\dagger R_{\rm out} =
R_{\rm in}^\dagger R_{\rm in} + X^\dagger X$.
(The input matrix \tt{X} is destroyed in the process.)  This is equivalent to the QR
definition of the update described above.

So the entire process might be coded using TMV as:
\begin{tmvcode}
int n_full = nrows_for_full_A_matrix;
int n_mem = nrows_that_fit_in_memory;
assert(n_mem <= n_full);
assert(n_mem > ncols);

tmv::Matrix<double> A(n_mem,ncols); 
tmv::Vector<double> b(n_mem);

// Import_Ab sets A to the first n_mem rows of the full matrix, 
// and also sets b to the same components of the full rhs vector.
// Maybe it reads from a file, or performs a calculation, etc.
Import_Ab(0,n_mem,A,b);

// x will be the solution to A_full x = b_full when we are done
// But for now, it is accumulating A_full.transpose() * b_full.
tmv::Vector<double> x = A.transpose() * b;

// Do the initial QR decomposition:
QR_Decompose(A.view());
tmv::UpperTriMatrix<double> R = A.upperTri();

// Iterate until we have done all the rows
for(int n1=n_mem, n2=n1+n_mem; n2<n_full; n1=n2, n2+=n_mem) {
    if (n2 > n_full) n2 = n_full;

    // Import the next bit:
    Import_Ab(n1,n2,A,b);

    // (Usually, A1==A, b1==b, but not the last time through the loop.)
    tmv::MatrixView<double> A1 = A.rowRange(0,n2-n1);
    tmv::VectorView<double> b1 = b.subVector(0,n2-n1);

    // Update, x, R:
    x += A1.transpose() * b1;
    QR_Update(R,A1);
}

// Finish the solution:
x /= R.transpose();
x /= R;
\end{tmvcode}

If the update only needs to add a single row, you can also use the form
\begin{tmvcode}
void QR_Update(UpperTriMatrix<T>& R, Vector<T>& v)
\end{tmvcode}
where \tt{v} is the single row that you want to have added to the matrix.  Again, \tt{v} is destroyed in the process.

\subsection{Downdate a QR decomposition}
\index{QR decomposition!Downdate}
\label{QRDowndate}

When performing a least-square fit of some data to a model,
it is common to do some kind of outlier rejection to remove data that
seem not to be applicable to the model - things like spurious measurements
and such.
For this, we basically want the opposite of a QR update - instead we want to 
find the QR decomposition that results from
removing a few rows from $A$.  This is called a QR ``downdate'', and is performed
using the subroutine:
\begin{tmvcode}
void QR_Downdate(UpperTriMatrix<T>& R, Matrix<T>& X)
\end{tmvcode}
where \tt{X} represents the rows from the original matrix to remove from the 
QR decomposition.  (The input matrix \tt{X} is destroyed in the process.)  If you only want to
remove a single row, you can also use
\begin{tmvcode}
void QR_Downdate(UpperTriMatrix<T>& R, Vector<T>& v)
\end{tmvcode}

It is possible for the downdate to fail (and throw an exception) 
if the matrix $X$ does not represent rows
of the matrix that was originally used to create $R$.
Furthermore,
with round-off errors, the error may still result with actual rows from the 
original $A$
if $R$ gets too close to singular.  In this case, \tt{QR\_Downdate} throws
a \tt{NonPosDef} exception.  This might seem like a strange choice, but the 
logic is that $R^\dagger R$ is the Cholesky decomposition of $A^\dagger A$,
and \tt{QR\_Downdate(R,X)} basically updates $R$ to be the Cholesky decomposition
of $A^\dagger A - X^\dagger X$.  The procedure fails (and throws) when this latter 
matrix is found not to be positive definite.
\index{QR decomposition!Downdate!NonPosDef exception}
\index{Exceptions!NonPosDef}

It is worth pointing out that the algorithm used in TMV is a new one developed by
Mike Jarvis.  Most of the texts and online resources that discuss the 
QR downdate algorithm only explain how to do one row at a time, using a 
modification of the QR update using Givens rotations.  
If you are doing many rows, it is common that roundoff errors in such a 
procedure accumulate sufficiently for the routine to fail.  The TMV algorithm
instead downdates all of the rows together using a modification of the 
Householder reflection algorithm for updates.  This algorithm seems to be
much more stable than ones that use Givens rotations.  

The only references to a similar algorithm that I could find in the literature is 
described in the paper,
"Stability Analysis of a Householder-based Algorithm for Downdating the Cholesky Factorization", Bojanczyk and Steinhardt, 1991, Siam J. Sci. Stat. Comput. 12, 6, 
1255\footnote{
It seems that this paper has become a bit forgotten.  A recent paper,
``Efficient Algorithms for Block Downdating of Least Squares Solutions'', Yanev and Kontoghiorghes, 2004, Applied Numerical Mathematics, 49, 3, evaluates
five algorithms for doing the downdate.  However, all
of them are block versions of the Given matrix approach.  They do not consider 
any algorithms that use Householder matrices to do the downdate, and do not reference
the above paper by Bojanczyk and Steinhardt.  In addition, none of the papers that
do cite the Bojanczyk and Steinhardt paper seem to be about the general problem 
of QR (or Cholesky) downdating.
}.
This paper describes a similar algorithm to compute the downdated $R$ matrix
using Householder matrices.  However, the details of the computation are somewhat
different from the TMV algorithm.  Also, they only consider real matrices, and they 
do not include the block-householder techniques in their description to employ more
so-called ``level-3'' matrix operations.

Therefore, I will describe the TMV downdate algorithm here.  I think it is clearer to 
begin by describing the update algorithm in \S\ref{QRUpdate_Algorithm}, since it 
is quite similar to the algorithm we use for downdating, but is a bit easier to 
understand.  Then the downdate algorithm
is described in \S\ref{QRDowndate_Algorithm}.

\subsubsection{The update algorithm}
\index{QR decomposition!Update!algorithm}
\label{QRUpdate_Algorithm}

First lets look at the Householder algorithm for QR update:

Given the initial decomposition $A_0 = Q_0 R_0$, we want to find $R$ such that
\begin{align*}
A_1 = \left(\begin{array}{c}A_0 \\ X \end{array}\right) &= Q_1 R_1 \\
\left(\begin{array}{c}Q_0 R_0 \\ X \end{array}\right) &= Q_1 R_1 \\
\left(\begin{array}{cc}Q_0 & 0 \\ 0 & 1 \end{array}\right) 
\left(\begin{array}{c}R_0 \\ X \end{array}\right) &= Q_1 R_1 
\end{align*}
So if we perform a QR decomposition: 
\begin{equation*}
S \equiv \left(\begin{array}{c}R_0 \\ X \end{array}\right) = Q_S R_S
\end{equation*}
Then this is the $R$ we want: $R_1 = R_S$, and 
\begin{equation*}
Q_1 = \left(\begin{array}{cc}Q_0 & 0 \\ 0 & 1 \end{array}\right)  Q_S
\end{equation*}

For the following discussion, let $N$ be the number of rows (and columns) in $R_0$,
and $M$ be the number of rows in $X$.

To perform the decomposition, we multiply $S$ by a series of Householder reflections
on the left to zero out each column of $X$ one at a time.  Householder reflections 
are unitary, so their product in the reverse order is $Q_S$:
\begin{align*}
\left(\begin{array}{c}R_1 \\ 0 \end{array}\right) &=
  H_N H_{N-1} ... H_2 H_1 
  \left(\begin{array}{c}R_0 \\ A_1 \end{array}\right)  \\
Q_S &= H_1^\dagger H_2^\dagger ... H_{N-1}^\dagger H_N^\dagger 
\end{align*}

Householder reflections are defined as $H = I - \beta (x - y e_1)  (x - y e_1)^\dagger$
where $x$ is a (column) vector, $y$ is a scalar with $|y| = ||x||_2$, $e_1$ is the
basis vector whose only non-zero element is the first: $e_1(1) = 1$, and
$\beta = (||x||_2^2 - y^* x(1))^{-1}$.
They have the useful properties that $H x = y e_1$ and they are unitary:
$H^\dagger H = H H^\dagger = I$.
Furthermore, if $\beta$ is real, they are also hermitian: $H = H^\dagger$.

$H_1$ is defined for the vector 
$x = (R_0(1,1), 0, 0, ... , 0, 0, X(1,1), X(2,1), ... , X(M,1) )$ 
where the stretch of $0$'s includes a total of $(N-1)$ $0$'s. This value of 
$x$ completely determines the Householder matrix $H_1$ up to an arbitrary sign
on either $y$ or $\beta$ (or in general an arbitrary factor $e^{i \theta}$) which is 
chosen to minimize rounding errors.  The optimal choice is to choose
$y = -||x||_2\:x(1)/|x(1)|$, which makes $\beta$ real.  
However, the LAPACK choice is $y = -||x||_2 \:sign(real(x(1)))$, which means
$\beta$ is complex, and $H$ is not Hermitian\footnote{
This choice complicates a lot of the calling routines which use
Householder matrices, since you need to keep track of conjugation of the 
$\beta$ values.  Since TMV is designed to be able to call LAPACK
when possible, it is forced to follow the same convention.

In fact, it could be argued that the LAPACK convention is even ``wrong'' in the sense that
their Householder matrices are not actually ``reflections''.  A reflection is a 
unitary matrix whose determinant is $-1$.  The determinant of a Householder 
matrix as defined here is $-\beta^2/|\beta|^2$ which is $-1$ for real $\beta$, 
but not for complex $\beta$.  But we are stuck with their choice, so we allow $\beta$
to be complex in this discussion.
}.
\index{LAPACK!Householder matrices}

The product $H_1 S$ ``reflects'' the first column
of $X$ into the first diagonal element of $R_0$.  Because of all the $0$'s, 
most of $R_0$ is unaffected -- only the first row of $R_0$ and the rest of $X$
are changed.
The subsequent Householder reflections are defined similarly, each zeroing out
a column of $X$, and modifying the corresponding row of $R_0$ and the 
remaining elements of $X$.

At the end of this procedure, the matrix $R_0$ will be changed into the 
matrix $R_1$.  If desired, $Q_S$ (and then $Q_1$) 
may also be calculated in the process, but the 
TMV implementation of the QR update does not calculate $Q_1$.
If there is a demand for such a routine, it would not be hard to add it, 
but I think most applications of the update do not use the $Q$ matrix explicitly.

\subsubsection{The downdate algorithm}
\index{QR decomposition!Downdate!algorithm}
\label{QRDowndate_Algorithm}

Given the initial decomposition
\begin{equation*}
A_1 = \left(\begin{array}{c}A_0 \\ X \end{array}\right) = Q_1 R_1 
\end{equation*}
we want to find $R_0$ such that $A_0 = Q_0 R_0$.

The TMV algorithm to do this essentially performs the same steps as in the update
algorithm above,
but instead removes the effect of each $H$ from $R_1$.
This is easy to
do if we can determine what each $H$ is, since $H^{-1} = H^\dagger$, so we just
apply $H^\dagger$ to update each row of $R_1$.  The $X$ update takes
the regular $H$ matrix, since we need to replicate the steps that we would do
for an update to keep finding the correct values for the remaining columns of $X$.

All of the values in the vector $x$ needed to define $H_1$ are given, except for the first,
$R_0(0,0)$.  But this is easy to calculate, since
\begin{equation*}
|R_1(0,0)|^2 = |R_0(0,0)|^2 + ||X(1:M,0)||_2^2
\end{equation*}
This determines the $x$ vector, which in turn defines $H_1$
(modulo an arbitrary sign, which again is chosen to minimize rounding errors).
Thus, we can calculate $H_1$ and apply it as described above.  Each subsequent Householder
matrix is created and applied similarly for each column of $X$.  When we have finished
this process, we are left with $R_0$ in the place of $R_1$.

If at any point in the process, we find the calculated $|R_0(k,k)|^2 < 0$, then 
the algorithm fails.  In the TMV implementation, a \tt{NonPosDef} exception is thrown.

In practice, for both of these algorithms, we actually use a blocked implementation for updating
the $R$ and $X$ matrices.  We accumulate the effect of the Householder matrices until 
there are sufficiently many (e.g. 64), at which point we update the appropriate rows of the $R$
matrix and the rest of $X$.  Implementing this correctly is mostly a matter of keeping track
of which elements have been updated yet, making sure that whenever an element is used,
it is already updated, while delaying as much of the calculation as possible in order 
to make maximum
use of the so-called ``level-3'' matrix functions, which are the most efficient on modern computers.
We also make the additional improvement of using a recursive algorithm within each block,
which gains some additional level-3 operations, for a bit more efficiency.

